\PassOptionsToPackage{table}{xcolor}
\documentclass[justified,openany,nofonts]{tufte-book}
\usepackage{mooculus}

%\usepackage{showkeys} %% Useful for debugging

\setcounter{secnumdepth}{2}




% Prints the month name (e.g., January) and the year (e.g., 2008)
\newcommand{\monthyear}{%
  \ifcase\month\or January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or
  December\fi\space\number\year
}

% Generates the index
\usepackage{makeidx}
\makeindex

\newcommand{\xrefn}[1]{\ref{#1}}



\newenvironment{lemma}{\subsection*{Lemma}}{}
\newenvironment{remark}[1]{\subsection*{Remark: #1}}{}


%\def\exam{\null}
\def\pagerdef{\null}

\def\dfont{\bf}
\def\em{\it}           % for emphasis

\newcommand{\ds}{\displaystyle}

\let\ssk\smallskip \let\msk\medskip \let\bsk\bigskip

\usepackage{multicol}
\def\twocol{\begin{multicols}{2}}
\def\endtwocol{\end{multicols}}

\title{Calculus}
%\author{Jim Fowler and Bart Snapp}
\publisher{This document was typeset on \today.}
%\newcommand{\blankpage}{\newpage\hbox{}\thispagestyle{empty}\newpage}

%% % Prints an epigraph and speaker in sans serif, all-caps type.
%% \newcommand{\openepigraph}[2]{%
%%   %\sffamily\fontsize{14}{16}\selectfont
%%   \begin{fullwidth}
%%   \sffamily\large
%%   \begin{doublespace}
%%   \noindent\allcaps{#1}\\% epigraph
%%   \noindent\allcaps{#2}% author
%%   \end{doublespace}
%%   \end{fullwidth}
%% }




\begin{document}
\maketitle

% v.4 copyright page


\begin{fullwidth}
~\vfill
\thispagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}
Copyright \copyright\ \the\year\ Jim Fowler and Bart Snapp

This work is licensed under the Creative Commons
Attribution-NonCommercial-ShareAlike License. To view a copy of this
license, visit
\url{http://creativecommons.org/licenses/by-nc-sa/3.0/}~or send a~or send a
letter to Creative Commons, 543 Howard Street, 5th Floor, San
Francisco, California, 94105, USA. If you distribute this work or a
derivative, include the history of the document. 

The source code is available
at: \url{https://github.com/ASCTech/mooculus/tree/master/public/textbook}

\noindent
This text is based on David Guichard's open-source calculus text which
in turn is a modification and expansion of notes written by Neal
Koblitz at the University of Washington. David Guichard's text is
available at \url{http://www.whitman.edu/mathematics/calculus/}~under a Creative Commons license.

\noindent The book includes some exercises and examples from {\it
  Elementary Calculus: An Approach Using Infinitesimals}, by H.~Jerome
Keisler, available at
\url{http://www.math.wisc.edu/~keisler/calc.html}~under a Creative
Commons license. 


\noindent This book is typeset in the Kerkis font, 
Kerkis \copyright~Department of Mathematics, University of the Aegean.


\noindent We will be glad to receive corrections and suggestions for
improvement at \texttt{fowler@math.osu.edu} or
\texttt{snapp@math.osu.edu}.

\end{fullwidth}


\tableofcontents


%% \renewcommand{\listtheoremname}{List of Main Theorems}
%% \setcounter{tocdepth}{1}
%% \listoftheorems[numwidth=4em,ignoreall,show={mainTheorem}]



\chapter*{List of Main Theorems}% uses ntheorem
\theoremlisttype{opt} 
\listtheorems{mainTheorem}



\chapter*{How to Read Mathematics}

Reading mathematics is \textbf{not} the same as reading a
novel. To read mathematics you need: 
\begin{enumerate}
\item A pen. 
\item Plenty of blank paper. 
\item A willingness to write things down.
\end{enumerate}
As you read mathematics, you must work along side of the text
itself. You must \textbf{write} down each expression, \textbf{sketch}
each graph, and \textbf{think} about what you are doing. You should
work examples and fill-in the details. This is not an easy task, it is
in fact \textbf{hard} work. However, mathematics is not a passive
endeavor. You, the reader, must become a doer of mathematics.



\setcounter{chapter}{-1}

%%
% Start the main matter (normal chapters)
%\mainmatter
%\tikzexternaldisable
\input functions
%\input limits
%\input infinityAndContinuity
%\input basicsOfDerivatives
%\input curveSketching
%\input productRuleAndQuotientRule
%\input chainRule
%\input derivativesOfTrigFunctions
%\input applicationsOfDifferentiation
%\input optimization
%\input linearApproximation
%\input antiderivatives
%\input integrals
%\input fundamentalTheoremOfCalculus
%\input techniquesOfIntegration
%\input applicationsOfIntegration



%\bibliography{sample-handout}
%\bibliographystyle{plainnat}

\finalizeanswers
\chapter*{Answers to Exercises}
\small
\addcontentsline{toc}{chapter}{Answers to Exercises}
\IfFileExists{answers.tex}{\input{answers}}
\normalsize
\backmatter
%\addcontentsline{toc}{chapter}{Index}
\printindex


\end{document}


