\documentclass[12pt]{amsart}
\usepackage{hyperref}
\newcommand{\mooculus}{\textsf{\textbf{mooc}ulus}}
\usepackage{nopageno}
\usepackage{geometry}
\geometry{left=1.5in,right=1.5in,top=1in,bottom=1in}

\title{What is Mooculus?}
\author{OSURA Dinner Series Presentation}

\begin{document}

\maketitle
\pagestyle{empty}
\thispagestyle{empty} 

On January~7, 2013, Ohio State launched its first massive open online
course (MOOC); the course is called ``Calculus One'' and is designed
to cover the same content as our local, in-person sections of
Math~1151.  Our online calculus course is available on Coursera at
\begin{center}
\texttt{https://www.coursera.org/course/calc1}
\end{center}
but since Coursera's platform lacks, for instance, randomly generated
math problems, we built our own MOOC platform here at Ohio State---we
call it \mooculus.  We encourage you to explore the platform we've
built by going to
\begin{center}
\texttt{https://mooculus.osu.edu/}
\end{center}

\subsection*{How many people are taking the course?}

The course has been offered twice: there were 47,227 students enrolled
in the first run, and there are 37,475 enrolled now in its second run.

%There are 32,890 students enrolled; a popular concern about MOOCs is
%the high attrition rate, but we have had 11,133 students engage with
%the course during Week~4, which is quite a healthy showing.  People
%are spending upwards of ten hours a week on our content.

\subsection*{Who are these students?}

It is easier to say who these students are not: they aren't the
traditional 18--22 year old college student.  Many of our MOOC
students already have advanced degrees.  A five year old took Calculus
One.  An 11-year-old beat her father on the final exam.  For others,
they took calculus decades ago (even $\geq 50$ years ago) and want a
refresher.  Some are in the hospital.  Some are soldiers, currently
deployed.  It's remarkably international and incredibly diverse.

\subsection*{How has the course been received?}

Remarkably well; posts on the forum include ``the teaching style is
the best I've seen on Coursera so far'' and ``one of the best
lecturers I've ever seen anywhere, live or online'' and ``this is the
first example I have seen in either Edx or Coursera of using the
medium really well.''  Thousands of people from all over the world are
successfully learning calculus with OSU's help.

\subsection*{Can students get credit for Calculus One?}

Not yet.  Right now, students who complete our MOOC will receive a
certificate signed by the instructor.  In the longer term, we hope to
be able to offer a ``signature track'' version of Calculus One.  With
signature track, a student's identity is verified, and the certificate
they earn will be verifiable and shareable, making them more valuable
as part of student's job or school application.

\pagebreak

\subsection*{How does a student engage with our course?}

For the student, \mooculus\ provides new content each week, in the form of
\begin{itemize}
\item another chapter to our free, open-source calculus textbook;
\item a dozen short lecture videos, between 2 and 15 minutes long; and
\item interactive randomly-generated exercises that emphasize both conceptual and computational learning.
\end{itemize}
Since one learns math by doing math, the interactive exercises are
especially important.  Unlike any other MOOC platform, we use
a Hidden Markov Model to estimate student learning, based on how the
student engages with our website.  Once \mooculus\ believes the
student has mastered a particular problem, the
student is provided with another challenge.

\subsection*{Why build our own platform?}

By building \mooculus, we provide a better student experience than any
other MOOC platform.  For instance, a common complaint with other
systems that don't rely on randomly generated exercises is that a
student eventually ``runs out'' of new exercises to try---\mooculus\
never runs out of problems.  Additionally, now that we've built
\mooculus, we can use the same platform for other courses, such as the
English department's WexMOOC.

\subsection*{MOOCs help high school teachers ``flip'' the classroom}

A ``flipped classroom'' is a classroom structured so that students
learn new content online through video lectures---watched in their
dorm room or at home, rather than during the assigned lecture
time---and what used to be homework is done in class.  The upside is
that this provides additional time for the teacher to interact with
students.  The downside is that teachers then have to prepare video
lectures, and shooting video is a very time-consuming process.

The content on \mooculus\ can help.  Currently we have six high school
teachers planning to use MOOCulus in conjunction with their AP
Calculus classes.  One measure of success will be the number of
teachers who use our platform.

\subsection*{MOOCs enhance the in-class experience at OSU}

While some believe that MOOCs are a replacement for in-class experiences, we do
not think this needs to be the case. Like Elizabeth Miller's project
\textit{Flipped and Flexible}, we see MOOCs as a tool for the
instructor to use their abilities to improve education across the
board. It should not be the case that the instructor is ``bound'' to
the textbook, instead the course materials should be dynamic enough
that the instructor can pick-and-choose what online materials are used
and even add to the materials at will.

\subsection*{What is the near-term future for MOOCs at OSU?}

We will be offering more courses, like Calculus Two and Complex
Analysis.  And we will be building a new platform: the Division of
Undergraduate Education at the NSF has completed processing a
recommendation for a Transforming Undergraduate Education in STEM
(TUES) Type 1 award for proposal DUE--1245433 (``Interactive
Textbook'') of \$180k, with a start date of September 1, 2013.  We
will be using that grant to build more online, interactive content.


\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
